Mechanical Detection of Dipole-Dipole Interactions \\between Electronic Spins in a Solid

### 2020-10-28 Mes papiers/Méca CR/Meca_cross.tex
confinement?

\author{C. Pellet-Mary, P. Huillery, M. Perdriat, G. H\'etet}

\affiliation{$^1$Laboratoire Pierre Aigrain, Ecole normale sup\'erieure, PSL Research University, CNRS, Universit\'e Pierre et Marie Curie, Sorbonne Universit\'es, Universit\'e Paris Diderot, Sorbonne Paris-Cit\'e, 24 rue Lhomond, 75231 Paris Cedex 05, France.}

@@

\begin{figure}[!ht]
  \centering \scalebox{0.45}{\includegraphics{Adamas_fixe.eps}}
  \caption{a) Photoluminescence signal in function of the scanning magnetic field. b) Predicted transition frequencies for the four classes of NV centers (dashed). ESR spectra for particular value of the magnetic field (plain, vertical)}
  \label{CR_deposited}
\end{figure}

\begin{figure}[!ht]
  \centering \scalebox{0.5}{\includegraphics{T1.eps}}
  \caption{Longitudinal relaxation from a single class of NV centers with exponential fit. i) The class is not at resonance with any other classes : T$_1$=1.02 ms. ii) The class is at resonance with another class : T$_1$=0.28 ms. iii) The class is at resonance with the three other classes : T$_1$=0.15 ms.
  }\label{T_1}
\end{figure}

@@

The diamonds that we use are in the form of a powder of particles that have a diameter of 15 $\mu$m. They are supplied by the company Adamas, which produces diamonds with a concentration of NV centers in the 3-4 ppm range. 
We operate with a Paul trap that is similar to the one used in \cite{delordPRL} with particles are stably trapped at the bottleneck region, where both the electric field gradient and anisotropy are stronger.
The microwave is applied directly on the trapping electrode, which provides an efficient means to excite the spins.
The photoluminescence is detected using standard confocal microscopy. We use about 100 $\mu$W of laser light at 532nm to polarise the NV centers. The laser is focused via a lens inside the vacuum chamber which has a numerical aperture of 0.5 and a working distance of 8mm. The focal point of the laser is kept few tens of micrometer away from the micro-diamond to mitigate the effect of radiation pressure and to enable laser excitation of the whole diamond \cite{delordPRL,delord2016}.

Let us first show that dipolar interaction in our diamonds is effective by studying cross-relaxation with diamonds that are simply attached to the trap.
In diamonds that contain more than 1 ppm of nitrogen-vacancy centers cross relaxation effects have been observed by many groups \cite{van_oort_cross-relaxation_1989, armstrong_nvnv_2010, jarmola_longitudinal_2015, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, holliday_optical_1989, mrozek_longitudinal_2015, jarmola_temperature-_2012, choi_depolarization_2017}.
It manifests itself as a change in the longitudinal relaxation rate of NV centers, which can then translate as a reduction of the overall photo-luminescence.

This effect has been routinely observed in bulk materials, the origin of which was attributed to the presence of charge tunneling amongst closely packed NV centers \cite{choi_depolarization_2017}. The NV centers that undergo tunneling with other impurities (possibly the substitutional nitrogen defect \cite{Manson}) have a reduced longitudinal spin lifetime $T_1$. When their energy matches the energy of NV centers that have phonon-limited lifetimes ($T_1\approx 5 ms$), the latter can drop 
to a few hundreds of micro-seconds \cite{Jarmola} when the NV concentration is large enough (typically in the ppm range)
This process has recently revived interest recently because the sensitivity of magnetometers is ultimately limited by these interactions \cite{Zhou}. 
%In our spin-mechanics experiment, we use micro-diamond particules subjected to a green laser field and a magnetic field.
%To observe the dipolar effects in a levitating platform, we employ heavily doped micro-diamonds.
In our experiment, we first probe CR by using a fixed bias magnetic field $\bf B_{\rm bias} \approx$ 100 G and scan another magnetic field $\bf B_{\rm em}$ at some angle with $B$ using an electromagnet. 
Figure \ref{CR_deposited}-a) shows the photoluminescence from the NV centers as a function of $\bf B_{\rm em}$. 
For these measurement we use a 1 mW green laser excitation which yields a repolarisation rate of about 100 $\mu$s. 
Three dips in the fluorescence can be observed. 
We attribute these to cross-relaxation that takes place when two orientation meet. To be sure, we recorded Electronic spin resonance spectra for different values which allows us to measure the total magnetic field felt by the spins, and therefore to extrapolate the transition frequencies of the four classes of NV for any applied $\bf B_{\rm em}$.
There is indeed a strong correspondence between the PL dips and the degeneracy conditions. 
(See SI for details on planes that cross etc.). 
We tentatively attribute those to CR. 
%It is important to characterize this effect further when using smaller particles such micro-diamonds where surface effects can play a role. 
%where, to our knowledge, such cross-relaxations have not been observed. 
%Indeed, surface effects may start to play a role and change quantitatively. 
To confirm our interpretation, and to extract the spin-torque dynamics in the mechanical experiment, we measure the longitudinal relaxation time ($T_1$ time) of the NV spins under different degeneracy conditions.

The measurement consists in applying a green laser to polarize the NV centers and to measure the photoluminescence at a later time. 
Such a measurement turns out to be hidden by recharging of NV centers in the dark, as has been observed and analyzed by several groups \cite{choi_depolarization_2017, mrozek_longitudinal_2015, giri_selective_2019, giri_coupled_2018}.
In order to accurately measure the $T_1$ and remove the changing PL due to the recharging effects, we use a sequence presented in the SI, where a microwave pulse is applied or not prior to relaxation. Subtracting the two resulting signals eliminates charge dynamics from the spin decay.
The results of the measurements are shown in Fig. \ref{T_1} for different degeneracies. 
It can be seen that $T_1$ is greatly shortened, from X to Y when approaching degeneracy, confirming this interpretation.

@@

One major ingredient of the present study compared to \cite{DelordNat} is the spin-torque coming from the NV centers in the absence of the microwave tone.
For this, let us consider first the torque coming from a single spin due to the transverse component of the magnetic field when the NV is in the ground state.
The hamiltonian for one NV orientation with quantization axis $z$ reads 
\begin{equation}\hat{H}_{\rm NV}=\hbar D \hat{S}_z^2+ \hbar \gamma_e \bf B  \cdot \bf\hat S.
\end{equation}
Under the condition $\gamma B \ll D$, $H_{B}= \gamma_e B  ( S_x \cos\theta + S_z \sin\theta)$ can be treated as a perturbation to the anisotropic part of the hamiltonian. 
The perturbed energy $\epsilon_g$ of $\ket{0}$ due to the transverse B field is then
\begin{equation} \epsilon_g=\sum_{m_s=\pm 1}\frac{ |\bra{0} H_B \ket{\pm 1}|^2}{-\epsilon^0_{\pm 1}}=-\hbar\frac{(\gamma_e B)^2 }{D} \cos^2\theta.
\end{equation}
One can deduce the total torque when N spins of a given orientation are polarised in the ground state to be 
\begin{equation}\tau_g=-N \frac{\partial \epsilon_0}{\partial \theta}=\hbar N  \frac{(\gamma_e B)^2}{D} \sin 2\theta.\end{equation}

At an angle $\theta=\pi/4$, where the torque is maximized, using $10^9$ spins polarized in the ground state and a B field of 100 G, the static torque is about $10^{-19}$ N.m.
We checked numerically that at these magnetic field values and angle, the NV are still polarized mostly in the ground state. 
Note that the torque is zero at $\theta=mod (\pi/2)$ where the eigenstates of $S_z$ are not mixed. The hamiltonian ground state is the eigenstate $| S=1,m_s=0\rangle$, which is is non magnetic. 
%Note also that the torque tends to zero if  tends to zero at any angle $\theta$. This is the case in the presence of strong spin-depolarisation, i.e. when $\hat\rho\rightarrow\frac{1}{3}\hat{I}$ (with $\hat{I}$ the identity operator).

@@

Let us now explain why CR modifies the NV spin torque.
Fig. \ref{principle}-i) is a depiction of a levitating diamond containing NV centers that have four possible directions along the $<111>$ axes. 
In the presence of a magnetic field, each NV centers has a different magnetic field projection, which means a different torque. 
Below, we show the sum of the ground state spin-torques $\tau_{NR}$, resulting from the four classes. 
Fig. \ref{principle}-ii) shows the same situation, with a magnetic field that is tuned so that two NV classes are co-resonant. 
When the re-polarisation rate due to the green laser is smaller than the cross-relaxation rate, the magnetization coming from these two classes (indicated by a circle) tends to cancel. The population in the three NV eigenstates will indeed equilibrate, giving a smaller magnetization.
The resulting torque is thus changed to $\tau_{R}$ and the diamond then turns about another angle.
This is the essence of the proposed dipolar detection.

@@

To observe this effect we use similar parameters and magnetic field arrangement than when the diamonds where outside the trap.
Let us first monitor the spin-mechanical resonances. 
Fig. \ref{data}-a) shows the signal coming from the laser reflected off the diamond surface as a function of the microwave signal driving the NV centers' spins, with a zoom on the $m_s=-1$ states frequencies.
Three ESR have been taken for three different $\bf B_{\rm em}$ amplitudes. 
At 10 and 25 G, one can observe 4 pics in the spectrum, that are demonstrate spin induced torque on the diamond from the 4 classes of NVs.
At 17 G however, two classes merge at 2.75 GHz. This is where we expect CR.

>>>>

\author{C. Pellet-Mary$^1$, P. Huillery$^1$, M. Perdriat$^1$, G. H\'etet}

\affiliation{Laboratoire De Physique de l'\'Ecole Normale Sup\'erieure, \'Ecole Normale Sup\'erieure, PSL Research University, CNRS, Sorbonne Universit\'e, Universit\'e de Paris , 24 rue Lhomond, 75231 Paris Cedex 05, France.}

@@

\begin{figure*}[!ht]
  \centering \scalebox{0.33}{\includegraphics{Adamas_fixe.eps}}
  \caption{Depolarization of NV ensembles induced by co-resonances. a) Representation of the NV-NV depolarization process. Green arrows represent the optical pumping in the brighter $\ket{m_s=0}$ state, curvy blue arrows represent the longitudinal relaxation where NV$_2$ has a considerably shorter T$_1$ than NV$_1$. Red circles represent the population in each state and red dashed arrows represent the resonant dipole-dipole interaction between the two NV centers. b) Photoluminescence signal in function of the scanning magnetic field. c) Predicted transition frequencies for the four classes of NV centers (dashed). ESR spectra for particular value of the magnetic field (plain, vertical). d) Longitudinal relaxation from a single class of NV centers with exponential fit. i) The class is not at resonance with any other classes : T$_1$=1.02 ms. ii) The class is at resonance with another class : T$_1$=0.28 ms. iii) The class is at resonance with the three other classes : T$_1$=0.15 ms.}
  \label{CR_deposited}
\end{figure*}

@@

The diamonds that we use are in the form of a powder of particles that have a diameter of 15 $\mu$m. They are supplied by the company Adamas, which produces diamonds with a concentration of NV centers in the 3-4 ppm range. 
We operate with a Paul trap that is similar to the one used in \cite{delordPRL} with particles stably trapped at the bottleneck region, where both the electric field gradient and anisotropy are stronger.
The microwave is applied directly on the trapping electrode, which provides an efficient means to excite the spins.
The photoluminescence is detected using standard confocal microscopy. We use about 100 $\mu$W of laser light at 532nm to polarise the NV centers. The laser is focused via a lens inside the vacuum chamber which has a numerical aperture of 0.5 and a working distance of 8mm. The focal point of the laser is kept few tens of micrometer away from the micro-diamond to mitigate the effect of radiation pressure and to enable laser excitation of the whole diamond \cite{delordPRL,delord2016}.

Let us first show that dipolar interaction in our diamonds is effective by studying cross-relaxation with diamonds that are simply attached to the trap.
Fig. \ref{CR_deposited} a) shows the interaction between two NV centers, one of which (NV$_2$) has a considerably reduced relaxation time T$_1$, almost negating the optical polarization by the laser. When considering the resonant dipole-dipole interaction with another center NV$_1$ (whose T$_1$ is phonon-limited), the centers will exchange excitation quanta through flip-flop process, resulting in a depolarization of NV$_1$. Generalizing, this means that a few fast-decaying NV can depolarize an ensemble of "normal" NV centers through dipolar interaction, reducing the average T$_1$ of the ensemble to a few to a few hundreds of micro-seconds \cite{Jarmola} and lowering the total photoluminescence. 
This effect has been observed by many groups\cite{van_oort_cross-relaxation_1989, armstrong_nvnv_2010, jarmola_longitudinal_2015, akhmedzhanov_microwave-free_2017, akhmedzhanov_magnetometry_2019, holliday_optical_1989, mrozek_longitudinal_2015, jarmola_temperature-_2012, choi_depolarization_2017}.

This effect has been routinely observed in bulk materials with large enough  NV concentration, typically in the ppm range. The origin of it (the fast-decaying NV centers)  was attributed to the presence of charge tunneling amongst closely packed NV centers \cite{choi_depolarization_2017}. The NV centers that undergo tunneling with other impurities (possibly the substitutional nitrogen defect \cite{manson_nv_2018}) have a largely reduced longitudinal spin lifetime $T_1$.
This process has recently revived interest because the sensitivity of magnetometers is ultimately limited by these interactions \cite{Zhou}.

%In our spin-mechanics experiment, we use micro-diamond particules subjected to a green laser field and a magnetic field.
%To observe the dipolar effects in a levitating platform, we employ heavily doped micro-diamonds.
In our experiment, we first probe CR by using a fixed bias magnetic field $\bf B_{\rm bias} \approx$ 100 G and scan another magnetic field $\bf B_{\rm em}$ at some angle with $B$ using an electromagnet. 
Figure \ref{CR_deposited}-b) shows the photoluminescence from the NV centers as a function of $\bf B_{\rm em}$. 
For these measurement we use a 1 mW green laser excitation which yields a repolarisation rate of about 100 $\mu$s. 
Three dips in the fluorescence can be observed. 
We attribute these to cross-relaxation that takes place when two orientation meet. To be sure, we recorded Electronic spin resonance spectra for different field values in Fig. \ref{CR_deposited}-c) which allows us to measure the total magnetic field felt by the spins, and therefore to extrapolate the transition frequencies of the four classes of NV for any applied $\bf B_{\rm em}$.
There is indeed a strong correspondence between the PL dips and the degeneracy conditions. 
This is expected for NV ensembles presenting a NV-NV depolarization process, since when two orientations or more are at resonance, the effective density of NV being able to exchange spin quanta increases, resulting in a stronger depolarization process.
%It is important to characterize this effect further when using smaller particles such micro-diamonds where surface effects can play a role. 
%where, to our knowledge, such cross-relaxations have not been observed. 
%Indeed, surface effects may start to play a role and change quantitatively.

To confirm our interpretation, and to extract the spin-torque dynamics in the mechanical experiment, we measure the longitudinal relaxation time ($T_1$ time) of the NV spins under different degeneracy conditions. %Clément : je comprend pas la partie "to extract the spin-torque dynamics"
The measurement presented in Fig \ref{CR_deposited}-d) consists in applying a green laser to polarize the NV centers and to measure the photoluminescence at a later time. 
Such a measurement turns out to be hidden by recharging of NV centers in the dark, as has been observed and analyzed by several groups \cite{choi_depolarization_2017, mrozek_longitudinal_2015, giri_selective_2019, giri_coupled_2018}.
In order to accurately measure the $T_1$ and remove the changing PL due to the recharging effects, we use a sequence presented in the SI, where a microwave pulse is applied or not prior to relaxation. Subtracting the two resulting signals eliminates charge dynamics from the spin decay.
The results are shown for different degeneracy configuration, starting with a case without degeneracy where we extract a $T_1=1.08$ ms, already shorter than the expected phonon limited lifetime, and this lifetime is even further reduced to 0.28 ms (resp. 0.15 ms) when two (resp. four) orientations are brought to resonance.

@@

One major ingredient of the present study compared to \cite{DelordNat} is the spin-torque coming from the NV centers in the absence of the microwave tone.
For this, let us consider first the torque coming from a single spin due to the transverse component of the magnetic field when the NV is in the ground state.
The Hamiltonian for one NV orientation with quantization axis $z$ reads 
\begin{equation}\hat{H}_{\rm NV}=\hbar D \hat{S}_z^2+ \hbar \gamma_e \bf B  \cdot \bf\hat S.
\end{equation}
Under the condition $\gamma B \ll D$, $H_{B}= \gamma_e B  ( S_x \cos\theta + S_z \sin\theta)$ can be treated as a perturbation to the anisotropic part of the Hamiltonian. 
The perturbed energy $\epsilon_g$ of $\ket{0}$ due to the transverse B field is then
\begin{equation} \epsilon_g=\sum_{m_s=\pm 1}\frac{ |\bra{0} H_B \ket{\pm 1}|^2}{-\epsilon^0_{\pm 1}}=-\hbar\frac{(\gamma_e B)^2 }{D} \cos^2\theta.
\end{equation}
One can deduce the total torque when N spins of a given orientation are polarized in the ground state to be 
\begin{equation}\tau_g=-N \frac{\partial \epsilon_0}{\partial \theta}=\hbar N  \frac{(\gamma_e B)^2}{D} \sin 2\theta.\end{equation}

At an angle $\theta=\pi/4$, where the torque is maximized, using $10^9$ spins polarized in the ground state and a B field of 100 G, the static torque is about $10^{-19}$ N.m.
We checked numerically that at these magnetic field values and angle, the NV are still polarized mostly in the ground state. 
Note that the torque is zero at $\theta=mod (\pi/2)$ where the eigenstates of $S_z$ are not mixed. The Hamiltonian ground state is the eigenstate $| S=1,m_s=0\rangle$, which is is non magnetic. 
%Note also that the torque tends to zero if  tends to zero at any angle $\theta$. This is the case in the presence of strong spin-depolarisation, i.e. when $\hat\rho\rightarrow\frac{1}{3}\hat{I}$ (with $\hat{I}$ the identity operator).

@@

Let us now explain why CR modifies the NV spin torque.
Fig. \ref{principle}-i) is a depiction of a levitating diamond containing NV centers that have four possible directions along the $<111>$ axes. 
In the presence of a magnetic field, each NV centers has a different magnetic field projection, which means a different torque. 
Below, we show the sum of the ground state spin-torques $\tau_{NR}$, resulting from the four classes. 
Fig. \ref{principle}-ii) shows the same situation, with a magnetic field that is tuned so that two NV classes are co-resonant. 
When the re-polarization rate due to the green laser is smaller than the cross-relaxation rate, the magnetization coming from these two classes (indicated by a circle) tends to cancel. The population in the three NV eigenstates will indeed equilibrate, giving a smaller magnetization.
The resulting torque is thus changed to $\tau_{R}$ and the diamond then turns about another angle.
This is the essence of the proposed dipolar detection.

@@

To observe this effect we use similar parameters and magnetic field arrangement than when the diamonds where outside the trap.
Let us first monitor the spin-mechanical resonances. 
Fig. \ref{data}-a) shows the signal coming from the laser reflected off the diamond surface as a function of the microwave signal driving the NV centers' spins, with a zoom on the $m_s=-1$ states frequencies.
Three ESR have been taken for three different $\bf B_{\rm em}$ amplitudes. 
At 10 and 25 G, one can observe 4 peaks in the spectrum that demonstrate spin induced torque on the diamond from the 4 classes of NVs.
At 17 G however, two classes merge at 2.75 GHz. This is where we expect CR.

### 2020-10-28 Mes papiers/Méca CR/Meca_cross_SI.tex
confinement?

\section*{NV$^-$ center Theory}
\subsection{NV spin Hamiltonian}
The NV spin Hamiltonian in its fundamental level can be written as :
\begin{equation*}
  \hat{\mathcal{H}}_s=D S_z^2 + \gamma_e \textbf{B}\cdot\hat{\textbf{S}},
  \end{equation*} 
Where $D = (2\pi) 2.87$ GHz is the crystal field splitting and $\gamma_e = 28 $GHz/T is the electron gyromagnetic ratio. We neglected contributions of the strain and local electric field since we are working with magnetic fields of the order of 100G, as well as the hyper-fine interaction with $^{14}$N since we are working with ensembles with a typical inhomogeneous broadening $\frac{1}{T_2^*} = (2\pi) 5$ MHz.

@@

\subsection{Diamond crystalline axes and degeneracy conditions}
\begin{figure}[!ht]
  \centering \scalebox{0.45}{\includegraphics{SI_Cristallo}}
  \caption{Representation of the magnetic field in the diamond crystalline basis.}
	\label{cristallo}
\end{figure}
There are four possible crystalline axis for the NV centers (so-called "classes" of NV) which are depicted in Fig. \ref{cristallo} b) and correspond to the crystalline directions [$111$], [$1\bar 1 \bar 1$], [$\bar 1 1 \bar 1$] and [$\bar 1 \bar 1 1$].

@@

\section*{Experimental details}

@@

The experimental setup illustrated in Fig.\ref{Optics} is very similar to the one used in \citep{DelordPRL} with the addition of a permanent magnet and an electromagnetic (EM) coil in order to perform magnetic filed scans. The diamond sample is typically illuminated with 1mW of 532 nm laser light, focused by a NA = 0.5 objective. An acousto-optic modulator (AOM) is used to switch on and off the 532nm laser and to finely tuned its power. The photo-luminescence (PL) is collected by the objective, separated form the excitation light using a dichroic mirror (DM) and a 532nm notch filter (NF), and detected using a multimode-fibered single-photon avalanche photo-detector (APD) (SPCM-AQRH-15 from Perkin Elmer). Typically, we detect PL photons at a rate of 1MHz.

The Paul trap is a pseudo-ring as can be seen in \citep{DelordPhD} which acts both as trap through the high voltage (HV) and as a microwave (MW) antenna.

The magnetic field generated by the (homemade) EM coil is controlled by a frequency generator (HP 33120A) with a triangle signal in order to perform magnetic field ramps. For scans of less than 100 G, no amplification of the signal is required.

@@

\subsection{$T_1$ Measurement}
\begin{figure}[!ht]
  \centering \scalebox{1}{\includegraphics{T1_soustraction_shema}}
  \caption{T$_1$ measurement protocol for a single class in an ensemble of NV centers}
	\label{T1_protocol}
\end{figure}
The protocol to measure the spin lifetime of the NV centers, as presented in Fig 2. of the main text, is described in Fig. \ref{T1_protocol} . The spins are initially polarized in their $m_s=\ket{0}$ state through a 1 ms green laser excitation pulse and then left to evolve in the dark for a variable dark time $\tau$. The spin state is finally read out with a 10 $\mu$s laser pulse, shorter than the polarization time of the spins.

The sequence is repeated a second time, adding a resonant microwave $\pi$ pulse on a transition of one of the four classes of NV$^-$, exchanging the population of the $m_s=\ket{0}$ state with one of the $m_s=\ket{\pm 1}$ state.

The two sequences are then subtracted, allowing us to measure the spin state population of a single class, and removing unwanted contributions to the photoluminescence, such as charge state transfer in the dark.

@@

To calibrate the magnetic field magnitude $B$, and its orientation $\theta$ with respect to the NV axis, we record Optically Detected Magnetic Resonace (ODMR) spectra....

@@

\section*{Cross-relaxation detection for another type of degeneracy}

\section*{Simulation details}
In this part we will discuss the method used to simulate the average torque as well as the $\ket{m_s=0}$ population of the stationary state. The numerical solving of the master equations was performed thanks to Quantum Toolbox in Python (QuTiP) \citep{qutip1} \citep{qutip2}.

@@

We also introduce the usual T1 jump operators...

In order to describe the T1 modification induced by the cross-relaxations, we use a phenomenological model ...

According to previous measurements \citep{choi_depolarization_2017}, only the $\ket{0}\bra{\pm1}$ (corresponding to a single quantum exchange in the dipole-dipole interaction) operators are modified by the cross-relaxations.

\bibliography{trap}

>>>>

\section{NV$^-$ center Theory}
\subsection{NV spin Hamiltonian}
The NV spin Hamiltonian in its fundamental level can be written as :
\begin{equation*}
  \hat{\mathcal{H}}_s=D S_z^2 + \gamma_e \textbf{B}\cdot\hat{\textbf{S}},
  \end{equation*} 
Where $D = (2\pi) 2.87$ GHz is the crystal field splitting originating from a spin-spin interaction, and $\gamma_e = 28 $GHz/T is the electron gyromagnetic ratio. We neglected contributions of the strain and local electric field since we are working with magnetic fields of the order of 100G, as well as the hyper-fine interaction with $^{14}$N since we are working with ensembles with a typical inhomogeneous broadening $\frac{1}{T_2^*} \approx (2\pi) 5$ MHz.

@@

\subsection{Diamond crystalline axes and degeneracy conditions}
\begin{figure}[!ht]
  \centering \scalebox{0.45}{\includegraphics{SI_Cristallo}}
  \caption{a) Representation of the four crystalline axes of the diamond b) Representation of the magnetic field in the diamond crystalline basis c) Representation in the $(\theta , \phi)$ basis of the crystalline planes ([110]$_\perp$ family of planes in dashed lines, [100]$_\perp$ in plain lines), [100] direction as a red circles, [111] direction as a green squares and the path of the magnetic field in the experiment of Fig. 3 of the main text in blue dots.}
	\label{cristallo}
\end{figure}
There are four possible crystalline axis for the NV centers (so-called "classes" of NV) which are depicted in Fig. \ref{cristallo} b) and correspond to the crystalline directions [$111$], [$1\bar 1 \bar 1$], [$\bar 1 1 \bar 1$] and [$\bar 1 \bar 1 1$].

@@

\section{Experimental details}

@@

The experimental setup illustrated in Fig.\ref{Optics} is very similar to the one used in \citep{DelordPRL} with the addition of a permanent magnet and an electromagnetic (EM) coil in order to perform magnetic field scans. The diamond sample is typically illuminated with 1mW of 532 nm laser light, focused by a NA = 0.5 objective. An acousto-optic modulator (AOM) is used to switch on and off the 532nm laser and to finely tuned its power. The photo-luminescence (PL) is collected by the objective, separated form the excitation light using a dichroic mirror (DM) and a 532nm notch filter (NF), and detected using a multimode-fibered single-photon avalanche photo-detector (APD) (SPCM-AQRH-15 from Perkin Elmer). Typically, we detect PL photons at a rate of 1 MHz.

The Paul trap is a pseudo-ring with a diameter of approximately 200 $\mu$m, as can be seen in \citep{DelordPhD}. It acts both as trap through the high voltage (HV) and as a microwave (MW) antenna.

The magnetic field generated by the (homemade) EM coil is controlled by a programmable power supply (Rhode \& Shwartz NGE 103) performing current ramps.

@@

\subsection{$T_1$ Measurement}
\begin{figure}[!ht]
  \centering \scalebox{1.2}{\includegraphics{T1_soustraction_shema}}
  \caption{T$_1$ measurement protocol for a single class in an ensemble of NV centers. Green bars represent laser excitation, red bar represent microwave $\pi$ pulse for the selected class}
	\label{T1_protocol}
\end{figure}
The protocol to measure the spin lifetime of the NV centers, as presented in Fig 2. of the main text, is described in Fig. \ref{T1_protocol} . In the first sequence ($S_1$), the spins are initially polarized in their $m_s=\ket{0}$ state through a 1 ms green laser excitation pulse and then left to evolve in the dark for a variable dark time $\tau$. The spin state is finally read out with a 10 $\mu$s laser pulse, shorter than the polarization time of the spins.

The second sequence ($S_2$) uses the same parameters (polarization time, dark time and readout time) than the first sequence but adds a resonant microwave $\pi$ pulse on a transition of one of the four classes of NV$^-$ right before the readout pulse, exchanging the population of the $m_s=\ket{0}$ state with one of the $m_s=\ket{\pm 1}$ state.

By looking a the subtraction of both signal $S_1-S_2$, we can then measure the spin state population of a single class, and remove unwanted contributions to the photoluminescence, such as charge state transfer in the dark.

@@

To calibrate the magnetic field magnitude $B$, and its orientation $\theta$ with respect to the NV axis, we record Optically Detected Magnetic Resonance (ODMR) spectra....

@@

\section{Cross-relaxation detection for another type of degeneracy}

\begin{figure}[!ht]
  \centering \scalebox{0.32}{\includegraphics{CR_meca_22}}
  \caption{Mechanical detection of a degeneracy in a [100]$_\perp$ plane. a) Theoretical frequencies for the four $\ket{0} \to \ket{-1}$ transitions in dashed lines and mechanical ESR spectra vertically. b) Photoluminescence of the NV centers as a function of the scanning magnetic field (blue crosses) and gaussian fit (orange line). c) Simulated population in $\ket{m_s=0}$ for the stationary state with (plain) or without (dashed) taking into account the decrease of the T$_1$ induced by the cross-relaxations.
  d) Reflected signal off the diamond as a function of the scanning magnetic field (blue crosses) and double gaussian fit (orange line) e) Simulated torque applied by the spins on the diamond, with (plain) or without(dashed) taking into account the cross-relaxations.}
	\label{CR_22}
\end{figure}

Similarly to Fig. 3 of the main text, we managed to mechanically detect another type of NV-NV resonance. Fig. \ref{CR_22}a) shows the theoretical frequencies of the $\ket{0} \to \ket{-1}$ transitions for all four classes of NV and ESR spectra measured through the reflected laser light off the diamond for various magnetic field values. Unlike The experiment of the main text, this time all four classes of NV are at resonance with another class at B=23 G, telling us that we are crossing a [100]$_\perp$ plane instead of a [110]$_\perp$ one.

Fig. \ref{CR_22}b) shows the recorded photoluminescence of the NV centers during the magnetic field scan. As previously, we see a drop in the luminosity when the resonance occurs, that we fitted with a gaussian. The drop is slightly more pronounced in this case since all classes are depolarized, instead of only two in the previous experiment. This behavior is predicted by the simulation of the $\ket{m_s=0}$ population in the stationary state in Fig. \ref{CR_22}c).

Fig. \ref{CR_22}d) shows the signal of the reflected laser off the diamond, allowing us to measure angular displacement. This time we see a clear difference with the experiment of the main text, instead of a single drop centered on the resonance, we see two bumps on both sides of the resonance, fitted with two gaussians.
This is in accordance with the simulated torque in Fig. \ref{CR_22}e) where we can see a dispersive profile with an almost zero torque at the resonance. The reason we do not observe a change of sign in the experiment (with two positive bumps instead of a positive and a negative one) might be because of the strong non-linearity of our detection : if the signal initially correspond to a dark spot of the speckle, then a change in the motion of the diamond can only result in an increased signal. We also note a drop both in the predicted torque amplitude, and in the signal-to-noise ratio  of the angular measurement compared to the single degeneracy case, implying a weaker effect.

Our physical interpretation of the angular signal we observe, as well as the dispersive profile of the torque is that in this case, for symmetry reason, the magnetic torque generated by the four classes of NV is not modified when we are exactly at resonance, since all four classes are affected identically. What happens when you are near, but not at, resonance is that the classes will not be identically depolarized : looking at Fig. \ref{CR_22}a), we can see that the two classes of higher frequency are always slightly closer to each other than the two classes of lower frequency (we can see the slope of the frequencies being a bit smaller for the two upper classes), resulting in a bit more depolarization for these two classes, except when they are exactly at resonance. This interpretation accounts both for the overall shape of the responses, and the comparatively weakness of the phenomenon.

\section{Simulation details}
In this part we will discuss the method used to simulate the average torque as well as the $\ket{m_s=0}$ population of the stationary state. The numerical solving of the master equations was performed thanks to the Quantum Toolbox in Python (QuTiP) \citep{qutip1} \citep{qutip2}.

@@

We also introduce the usual T$_1$ jump operators $\mathcal{L}_i^j= \frac{1}{T_1}\ket{i}\bra{j}$ where $\ket{i,j}$=$\ket{0, \pm 1}$

In order to describe the T$_1$ modification induced by the cross-relaxations, we use a phenomenological model where each class as its own T$_1^i$ (i from 1 to 4) that depends on the energy levels of the other classes with the formula :
\begin{equation*}
\frac{1}{T_1^i}=\frac{1}{T_1^0}+\sum_{j \neq i} \frac{1}{T_1^{dd}}e^{-\frac{(\nu_i-\nu_j)^2}{2(\sigma^{dd})^2}},
\end{equation*}
where $\nu_i$ and $\nu_j$ are the transition frequencies of the classes i and j (we are arbitrarily considering the $\ket{0} \to \ket{-1}$ transition here, since the resonance condition is the same for both transitions at the magnetic fields we are working at. This would not always be true for magnetic fields greater than 592 G\citep{van_oort_cross-relaxation_1989}).

$\sigma^{dd} = 6$ MHz is the the dipolar interaction width, which we measured to be similar to the inhomogeneous broadening.

$T_1^0=1.03$ ms and $T_1^{dd}=0.38$ ms were chosen to match the T$_1$ measured in Fig. 1 of the main text. We only focus on the T$_1$ without degeneracy and the one with a single degeneracy since our experiments will not have more than two classes at resonance at once. Our model is probably not suited to deal with triple our quadruple resonances.

Finally, according to previous measurements \citep{choi_depolarization_2017}, only the $\ket{0}\bra{\pm1}$ and $\ket{\pm1}\bra{0}$ (corresponding to a single quantum exchange in the dipole-dipole interaction) operators are modified by the cross-relaxations.

With this model in consideration, we can numerically solve the master equation and get the density matrix in the stationary state $\rho_s$. With $\rho_s$ we can directly obtain the $\ket{m_s=0}$ population, corresponding experimentally to the photoluminescence.

As for the torque, we use a semi-classical formula :
\begin{equation*}
\mathbf \Gamma = N_0 \gamma_e <\hat{\mathbf S}> \times \mathbf B,
\end{equation*}
where $N_0 \approx 10^9$ is an estimate of the number of spins in our sample based on the average size and NV density of our diamonds, $\gamma_e$ is the gyromagnetic ratio of the electron and $<\hat{\mathbf S}> =\mathrm{Tr}(\rho_s \mathbf{\hat S})$ is the averaged spin vector in the stationary sate, averaged again on the four possible orientations of NV.
This formula assumes that the spin dynamics is faster than the dynamics of the motion of the diamond, which is the case in our experiments.

In our plots in Fig. X and Fig. 3 of the main ext, we only represent one spatial component (e.g. $\Gamma_x$) of the torque, but the three components behave similarly.
\bibliography{trap}
